Rotating Stars from \Kepler Observed with Gaia DR2

### 2018-05-09 paper/ms.tex
some words

The \Kepler mission \citep{borucki2010} a new era for enabling detailed study of angular momentum loss in low-mass stars. long noted as a means to possibly age-date stars (gyro), open clusters give hope that this general model works for low-mass stars
many Qs exist about details. these include Qs about initial rotation period distribution (CITE), the specific prescription for spin-down (ANGUS, etc), and exploring the efficiency of this angular momentum loss mechanism at old ages (Van Saders).

@@

In this letter we follow the work of \citet{davenport2017} in studying the \Kepler rotation period sample using astrometric data from the {\em Gaia} mission \citep{gaia}. By matching the \citet{mcquillan2014} rotation period catalog to the newest data from {\em Gaia} Data Release 2, we can use precise distances for essentially every star to select likely main sequence dwarfs to distances well over 1 kpc. Importantly this filters out both sub-giants (the main contaminant noted by \cite{davenport2017}), and unresolved binary stars.
Here we demonstrate the power of such a combined time-domain and astrometric sample for constraining the detailed evolution of main sequence stars themselves, and exploring the star formation history of the Milky Way.

%\citet{davenport2017} looked at this with Gaia DR1, extending period gap to G dwarfs, pointing to SFH as cause. sub giant contamination was the thing. Now we can finally follow up on this and study the full \Kepler rotation period sample

@@

data xmatched by M. Bedell between Kepler DR25 (CITE) and Gaia DR2 (CITE), using 1arcsec radius.
distances using the updated prescription from \citet{bailer-jones2018}

@@

\begin{figure*}[]
\centering
\includegraphics[width=6in]{../figures/cmd}
\caption{
Color--magnitude diagram for 30,305 \Kepler stars from the \citet{mcquillan2014} sample that are included in Gaia DR2, colored by their measured rotation period. For reference we show a $10^9$ year MIST isochrone (black line) used to select likely main sequence, single stars (black line). A track of binary  stars is apparent $\sim$0.7 mag above the main sequence. As in \citet{davenport2017}, we find significant contamination of the rotation period sample for bluer stars by sub-giants.}
\label{fig:cmd}
\end{figure*}

@@

This work made use of the \url{gaia-kepler.fun} crossmatch database created by Megan Bedell.

@@

%This research made use of the cross-match service, and the SIMBAD database, provided by CDS, Strasbourg, France.
%
%This work has made use of data from the European Space Agency (ESA)
%mission {\it Gaia} (\url{http://www.cosmos.esa.int/gaia}), processed by
%the {\it Gaia} Data Processing and Analysis Consortium (DPAC,
%\url{http://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding
%for the DPAC has been provided by national institutions, in particular
%the institutions participating in the {\it Gaia} Multilateral Agreement.

\bibliography{/Users/james/Dropbox/references}

>>>>

The \Kepler mission \citep{borucki2010} a new era for enabling detailed study of angular momentum loss in low-mass stars. long noted as a means to possibly age-date stars (gyro), open clusters give hope that this general model works for low-mass stars
many Qs exist about details. these include Qs about initial rotation period distribution \citep[e.g.][]{barnes2010,matt2015}, the specific prescription for spin-down \citep{angus2015}, and exploring the efficiency of this angular momentum loss mechanism at old ages \citep{van-saders2016}.

@@

In this letter we follow the work of \citet{davenport2017} in studying the \Kepler rotation period sample using astrometric data from the {\em Gaia} mission \citep{gaia}. By matching the \citet{mcquillan2014} rotation period catalog to the newest data from {\em Gaia} Data Release 2 \citep{gaia_dr2}, we can use precise distances for essentially every star to select likely main sequence dwarfs to distances well over 1 kpc. Importantly this filters out both sub-giants (the main contaminant noted by \cite{davenport2017}), and unresolved binary stars.
Here we demonstrate the power of such a combined time-domain and astrometric sample for constraining the detailed evolution of main sequence stars themselves, and exploring the star formation history of the Milky Way.

@@

Our 
data xmatched by M. Bedell between \Kepler and Gaia DR2 (CITE). using 1arcsec radius, which included 195,830 sources with stellar properties from the \Kepler Data Release 25 and {\em Gaia} DR2.

sub-set was matched to the rotation catalog of \citet{mcquillan2014}

To select stars with good parallaxes, as well as high quality photometry from {\em Gaia}, we selected stars with the following criteria:
\begin{itemize}
\item Parallax error $< 0.1$ mas
\item $\sigma(M_{G}) / M_{G} < 0.01$
\item $\sigma(G_{BP}) /G_{BP} < 0.01$
\item $\sigma(G_{RP}) /G_{RP} < 0.01$
\end{itemize}

distances using the updated prescription from \citet{bailer-jones2018}
Following the suggested use of \citet{bailer-jones2018}, we use only sources with {\tt modality\_flag} == 1 (i.e. not a bimodal distance solution) and {\tt result\_flag} == 1 (i.e. well constrained distance).

Our final sample contained 30,305 stars in Gaia DR2 with measured \Kepler rotation periods that passed these selection criteria. A color--magnitude diagram of this sample is presented in Figure \ref{fig:cmd}, with points colored by their rotation periods.

\begin{figure}[]
\centering
\includegraphics[width=3.6in]{../figures/cmd}
\caption{
Color--magnitude diagram for 30,305 \Kepler stars from the \citet{mcquillan2014} sample that are included in Gaia DR2, colored by their measured rotation period. For reference we show a $10^9$ year MIST isochrone (black line) used to select likely main sequence, single stars (black line). A track of binary  stars is apparent $\sim$0.75 mag above the main sequence. As in \citet{davenport2017}, we find significant contamination of the rotation period sample for bluer stars by sub-giants.}
\label{fig:cmd}
\end{figure}

@@

As in \citet{davenport2017}, the color--magnitude diagram shown in Figure \ref{fig:cmd} shows many of the bluer stars in the \citet{mcquillan2014} sample are significantly above the main sequence. These are likely subgiant stars, which do not follow the main sequence stars spin-down evolution \citep[e.g.][]{donascimento2012, van-saders2013}. Since \citet{davenport2017} found subgiants could obscure the rotation period bimodality for G dwarfs, these must be excluded from our analysis, but encourage future studies to explore the wealth of angular momentum evolution data from these most-main sequence objects.

Beyond the subgiant contamination above the main sequence, we can also see a secondary population of stars in a parallel track above the normal main sequence, which are likely unresolved binary stars. The pile-up above the main sequence occurs due to unresolved equal-mass field binaries, which was seen in the {\em Gaia} DR1 data as well \citep{anderson2017}. Since these systems may have experienced tidal evolution that could significantly impact their rotation evolution \citep[e.g.][]{lurie2017}, we must also remove these from our analysis. Though we do not explore the binary population in any detail here, this sample (perhaps with radial velocity follow-up) may provide useful insight into the tidal evolution of binary stars.

we use the Mesa Isochrones and Stellar Tracks \citep[MIST;][]{MIST} to describe the main sequence stars in Figure \ref{fig:cmd} using [Fe/H] =0.25, and an age of $10^9$ years.
main seq stars chosen in 0.5 mag window

@@

This work made use of the \url{gaia-kepler.fun} crossmatch database, created by Megan Bedell.

@@

This work has made use of data from the European Space Agency (ESA) mission
{\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it Gaia}
Data Processing and Analysis Consortium (DPAC,
\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding for the DPAC
has been provided by national institutions, in particular the institutions
participating in the {\it Gaia} Multilateral Agreement.

\bibliography{/Users/davenpj3/Dropbox/references}
